Integration of Multiple Constraints in ACG

### 2013-09-20 lenls/paper.tex
Expanding the Motivation and ACG sections

\mainmatter              % start of the contributions
%
\title{Integration of Multiple Constraints in ACG}
%
\titlerunning{Constraints in ACG}  % abbreviated title (for running head)
%                                    also used for the TOC unless
%                                    \toctitle is used
%
\author{Jiří Maršík\inst{1,2} \and Maxime Amblard\inst{2}}
%
\authorrunning{Maršík et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Jiří Maršík and Maxime Amblard}
%
\institute{Université de Lorraine, LORIA, Nancy, 54000, France\\
\email{\{jiri.marsik, maxime.amblard\}@loria.fr},\\
\and
INRIA, Nancy, 54000, France}

@@

Abstract Categorial Grammars (ACGs) have shown to be a viable formalism for
elegantly encoding the dynamic nature of discourse. Proposals based on
continuation semantics \cite{de2006towards} have tackled topics such as event
anaphora \cite{qian2011event}, SDRT discourse structure \cite{asher2011sdrt}
and modal accessibility constraints \cite{asher2011montagovian}. However, all
of these treatments only consider tiny fragments of languages. We are
interested in building a wide-coverage grammar which integrates and reconciles
the existing formal treatments of discourse and allows us to study their
interactions and to build discourse representations automatically.

In this paper, we focus on the problem of constructing a wide-coverage ACG. We
identify a problem that comes up when one tries to enforce multiple
constraints/dependencies in a single ACG and we propose a generalization of
the formalism which resolves the issue.

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/double-acg.pdf}
    \caption{\label{fig:acg-comp-basic} Connecting form with meaning.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/serial-over-parallel.pdf}
    \caption{\label{fig:acg-comp-constr} Adding a constraint.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/parallel-over-serial.pdf}
    \caption{\label{fig:acg-comp-sem} Distinguishing syntactic and
      semantic ambiguities.}
  \end{subfigure}
  \caption{\label{fig:acg-comp} Diagrams of systems of ACGs.}
\end{figure}

\section{Systems of ACGs}
\label{sec:acg-comp}

An ACG is defined in terms of two higher-order signatures, sets of typed
constants on which we can form well-typed lambda terms, and a
translating function from terms of one of the signatures to terms of the
other (the former signature being called the \emph{abstract signature},
the latter being called the \emph{object signature} and the function
being called a \emph{lexicon}). We are then interested in the terms of
the abstract signature having some distinguished type and their images
through the lexicon (these two sets of terms are called the
\emph{abstract language} and the \emph{object language}, respectively).

When describing languages, we are usually interested in systems of more than
one ACG, and diagrams such as the ones on Figure~\ref{fig:acg-comp} have been
used to explain them.

In Figure~\ref{fig:acg-comp-basic}, abstract syntactic terms from the
signature $\Sigma_{Synt}$ yield phonological strings and semantic
representations using two different lexicons. In
Figure~\ref{fig:acg-comp-constr}, this pattern is extended by adding a
new abstract signature which narrows the language of syntactic terms and
implements some linguistic constraint
\cite{pogodalla2012controlling}. Finally, in
Figure~\ref{fig:acg-comp-sem}, the abstract signature $\Sigma_{Synt}$ of
Figure~\ref{fig:acg-comp-basic} is split into two signatures,
$\Sigma_{Syntax}$ and $\Sigma_{SyntSem}$, in order to model syntactic
and semantic ambiguities in different grammars
\cite{pogodalla2007generalizing}.

>>>>

\mainmatter              % start of the contributions
%
\title{Integration of Multiple Constraints in ACG}
%
\titlerunning{Constraints in ACG}  % abbreviated title (for running head)
%                                    also used for the TOC unless
%                                    \toctitle is used
%
\author{Jiří Maršík\inst{1,2} \and Maxime Amblard\inst{1,2}}
%
\authorrunning{Maršík et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Jiří Maršík and Maxime Amblard}
%
\institute{Université de Lorraine, LORIA, Nancy, 54000, France\\
\email{\{jiri.marsik, maxime.amblard\}@loria.fr},\\
\and
INRIA, Nancy, 54000, France}

@@

Abstract Categorial Grammars (ACGs) \cite{de2001towards} have shown to be a
viable formalism for elegantly encoding the dynamic nature of
discourse. Proposals based on continuation semantics \cite{de2006towards} have
tackled topics such as event anaphora \cite{qian2011event}, SDRT discourse
structure \cite{asher2011sdrt} and modal accessibility constraints
\cite{asher2011montagovian}. However, all of these treatments only consider
tiny fragments of languages. We are interested in building a wide-coverage
grammar which integrates and reconciles the existing formal treatments of
discourse and allows us to study their interactions and to build discourse
representations automatically.

A presupposed condition to actually using an ACG to describe discourse in a
large scope is to have a large scale ACG grammar in the first place. The work
presented here was motivated by this quest for a wide-coverage ACG grammar, as
seen from a point of view of language as a system of largely orthogonal
constraints. Encoding ancillary constraints in a type system can make the
system hard to read and understand. If multiple constraints written in the
style of \cite{pogodalla2012controlling} are to be enforced in the same
grammar, we advocate for extending the formalism to prevent the incidental
complexity that would otherwise appear.

We begin with a short review of Abstract Categorial Grammars in which we
highlight the ways in which Abstract Categorial Grammars facilitate
decomposition of linguistic description. We follow that by a closer look at an
example linguistic constraint and how its treatment might look like in a
categorial grammar, which motivates our proposed extension. We then present
the Graphical Abstract Categorial Grammars and discuss some of its formal
properties. We then finish with an illustration of our approach which
incorporates linguistic contribution from multiple sources into a single
coherent and simple grammar and discuss the limitations and challenges of
adopting this approach.

\section{Abstract Categorial Grammars}
\label{sec:acg-comp}

We first review the grammatical framework in which we conduct our
work. Abstract Categorial Grammars are built upon two mathematical structures,
\emph{(higher-order) signatures} and \emph{lexicons}.

\subsection{Higher-Order Signatures}
\label{ssec:sig}

A \textbf{higher-order signature} is a set of elements that we call
\emph{constants}, each of which is associated with a type. Formally, it
is defined as a triple $\Sigma = \mathopen{<}A, C, \tau\mathclose{>}$,
where:
\begin{itemize}
  \item $C$ is the (finite) set of constants
  \item $A$ is a (finite) set of atomic types
  \item $\tau$ is the type-associating mapping from $C$ to
    $\mathcal{T}(A)$, the set of types built over $A$
\end{itemize}

In our case, $\mathcal{T}(A)$ is the implicative fragment of linear and
intuitionistic logic with $A$ being the atomic propositions. This means
that $\mathcal{T}(A)$ contains all the $a \in A$ and all the $\alpha \limp
\beta$ and $\alpha \to \beta$ for $\alpha, \beta \in \mathcal{T}(A)$.

A signature $\Sigma = \mathopen{<}A, C, \tau\mathclose{>}$, by itself, already
lets us define an interesting set of structures, that is the set
$\Lambda(\Sigma)$ of \emph{well-typed lambda terms} built upon the signature
$\Sigma$. To make this structure even more useful, we often focus ourselves
only on terms that have a specific \emph{distinguished type}. Using this
notion of a signature of typed constants and some distinguished type, we can
describe languages of, e.g. tree-like (and by extension string-like), lambda
terms.

\subsection{Lexicons}

The idea of a signature is coupled with the one of \textbf{lexicon},
which is a mapping between two different signatures (mapping the
constants of one into well-typed terms of the other). Formally speaking,
a lexicon $\mathcal{L}$ from a signature $\Sigma_1 = \mathopen{<}A_1,
C_1, \tau_1\mathclose{>}$ (which we call the abstract signature) to a
signature $\Sigma_2 = \mathopen{<}A_2, C_2, \tau_2\mathclose{>}$ (which
we call the object signature) is a pair $\mathopen{<}F, G\mathclose{>}$
such that:

\begin{itemize}
\item $G$ is a mapping from $C_1$ to $\Lambda(\Sigma_2)$ assigning to
  every constant of the abstract signature a term in the object
  signature, which can be understood as its
  interpretation/implementation/realization.
\item $F$ is a mapping from $A_1$ to $\mathcal{T}(A_2)$ which links the
  abstract-level types with the object-level types that they are
  realized as.
\item $F$ and $G$ are compatible, meaning that for any $c \in C_1$, we
  have $\vdash_{\Sigma_2} G(c) : \hat{F}(\tau_1(c))$ (we will be using
  $\hat{F}$ and $\hat{G}$ to refer to the homomorphic extensions of $F$
  and $G$ to $\mathcal{T}(A_1)$ and $\Lambda(\Sigma_1)$
  respectively).
\end{itemize}

Now we have enough machinery in hand to describe how ACGs define
languages. Given some signature and a distinguished type, we can talk of the
\emph{abstract language} which is the set of well-typed terms built on that
signature having that distinguished type. If we then also consider a lexicon
and its abstract signature, then the language we get by mapping the abstract
language of the signature through that lexicon is called an \emph{object
  language}.

An ACG for us will then be just a collection of signatures with their
distinguished types and the lexicons connecting these signatures, which we
will write down in diagrams such as the ones in Figure~\ref{fig:acg-comp}.

The most common pattern will have us using two object signatures for the
surface forms of utterances (strings) and their logical forms (logical
propositions) and an abstract signature which is connected to both of the
object signatures via lexicons (as we can see in
Figure~\ref{fig:acg-comp-basic}). Parsing is then just a matter of inverting
the surface lexicon to get the abstract term and then applying to it the
logical lexicon. Generation is symmetric, we simply invert the logical lexicon
and apply the surface lexicon.

\subsection{Systems of ACGs}

\begin{figure}[t]
  \centering
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/double-acg.pdf}
    \caption{\label{fig:acg-comp-basic} Connecting form with meaning.}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/serial-over-parallel.pdf}
    \caption{\label{fig:acg-comp-constr} Adding a constraint (Constraint).}
  \end{subfigure}
  \qquad
  \begin{subfigure}[b]{0.25\textwidth}
    \centering
    \includegraphics[height=0.2\textheight]{../diagrams/parallel-over-serial.pdf}
    \caption{\label{fig:acg-comp-sem} Distinguishing syntactic and
      semantic ambiguities (Elaboration).}
  \end{subfigure}
  \caption{\label{fig:acg-comp} Diagrams of systems of ACGs.}
\end{figure}

Besides the pattern in Figure~\ref{fig:acg-comp-basic}, there have been other
techniques of decomposing linguistic description in the ACG literature. Here
we will focus on two that are relevant to our work.

The first constitutes the Constraint pattern used throughout
\cite{pogodalla2012controlling}. Imagine we want our grammar to enforce a
particular linguistic constraint, say that tensed clauses form islands for
quantifier scope, and suppose we want to proceed by refining our types so as
to make linguistic structures violating the constraint untypable. We could
dive in to our grammar and rewrite it to use the new types. However, we can
also keep the existing grammar intact and complement it with a signature and a
lexicon that translates that signature to the abstract-most signature of our
existing grammar. Usually, the new signature will resemble the prior
abstract-most one, albeit with more fine-grained types necessary to express
the constraint in question.

This pattern is practical for examining linguistic constraints one-by-one and
demonstrating that the formalism is capable of expressing them
individually. However, this pattern does not serve us much in the context of
ACGs when we try to build a grammar which incorporates many of these at the
same time. This is due to the fact that after grafting on the first
constraint, its added signature now becomes the abstract-most signature in the
grammar. Any other constraint that would be added using the Constraint pattern
would have to be translated into this new signature. This would mean that the
type systems of newly added constraints would have to re-enforce all of the
constraints introduced before, therefore nullifying almost any benefit of the
pattern.

Another pattern we will briefly mention is the Elaboration pattern in which an
edge in the diagram (a lexicon) is expanded into two edges and an intermediate
node (signature). We can illustrate this pattern on Tree Adjoining Grammars
\cite{degroote02}. Suppose we have a signature describing TAG derivation trees
and a lexicon which maps them to their yield. We might want to make this
grammar richer, and arguably more synoptic, by elaborating the translation
from derivation trees to yields by introducing derived trees and describe
separately how both derivation trees produce derived trees and how derived
trees produce the yields. This way, we can make emerge a potentially useful
linguistic structure.

This pattern was used in \cite{pogodalla2007generalizing} to elaborate the
translation from the common Montague-style \cite{montague1973proper}
higher-order signature to its yield by introducing lower-order syntactic
structures more reminiscent of TAG or CFG trees. This helps make a clear
distinction between ambiguities caused by syntax and by semantics by
segregating the purely syntactic description in the $Syntax$ signature from
the issues of operator scope that have to be solved in the $SyntSem$ signature
(Figure~\ref{fig:acg-comp-sem}). However, in classical ACGs, we cannot exploit
this pattern as much as we would like. If we were to use this grammar as a
kernel on top of which we would like to express a syntactic constraint, we
would not be able to attach the constraint's signature directly to the
$Syntax$ node but we would instead need to attach it to
$SyntSem$\footnote{This is due to the fact that ACG diagrams are always
  arborescences, in which the root node represents the abstract language from
  whose terms are generated terms of all the object languages.}. This means
that the syntactic constraint would end up being expressed by strengthening
the type system of $SyntSem$, which only adds unnecessary complexity.
